---
title: "Capstone Analysis Project"
author: "Parker Reed"
date: "2025-04-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Process the Data
```{r}
library("tidyverse")
library("conflicted")

q1_2019<- read_csv("Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```
 Checking for inconsistencies in column naming
```{r}
colnames(q1_2019)
colnames(q1_2020)
```
#### Renaming Columns to Ensure Consistency
```{r}
(q1_2019 <- rename(q1_2019
  ,ride_id = trip_id
  ,rideable_type = bikeid
  ,started_at = start_time
  ,ended_at = end_time
  ,start_station_name = from_station_name
  ,start_station_id = from_station_id
  ,end_station_name = to_station_name
  ,end_station_id = to_station_id
  ,member_casual = usertype
  ))
```
 Checking and correcting data types so that the data frames are stackable
```{r}
str(q1_2019)
str(q1_2020)

q1_2019 <- mutate(q1_2019, ride_id = as.character(ride_id),
  rideable_type = as.character(rideable_type))
  
all_trips <- bind_rows(q1_2019, q1_2020)
  
```
 Removing variables that were not collected in both Q1 2019 and Q1 2020
```{r}
all_trips <- all_trips %>% 
  select(-c(start_lat,start_lng,end_lat,end_lng, birthyear, gender, tripduration))

```
## Preparing/Clean the Data for Analysis
```{r}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
str(all_trips)
summary(all_trips)
tail(all_trips)
head(all_trips)
glimpse(all_trips)

```
Based on the above, we can see at least three different distinct entries under the "member_casual" column, when we were only expecting two. 

```{r}
all_trips %>% 
    distinct(member_casual)
all_trips <- all_trips %>% 
    mutate(member_casual = recode(member_casual,
                                  "Subscriber" = "member",
                                  "Customer" = "casual"))
table(all_trips$member_casual)


```
Additionally, the date is in its long format, which means it can only be agggregated at ride level. 
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
We can also gain some useful insights if we knew how long each ride was: 
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
str(all_trips)
all_trips$ride_length<-as.numeric(as.character(all_trips$ride_length))
```
 There are certain entries in this data, where bikes were taken out for repair in which the ride_length would be 0
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
##Analyzing the Data
Generally descriptive analysis
```{r}
mean(all_trips_v2$ride_length)
median(all_trips_v2$ride_length)
max(all_trips_v2$ride_length)
min(all_trips_v2$ride_length)
```
Comparing members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN=mean)
aggregate(all_trips_v2$ride_length~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
Comparing members and casual users based on the day of the week. 
```{r}
all_trips_v2$day_of_week<-ordered(all_trips_v2$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
                                  
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
Analyzing ridership data by type and weekday. 
```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)
```

## Visualizing the Data

```{r pressure, echo=FALSE}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
    summarise(number_of_rides = n()
  ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(x="Day of the Week", y = "Number of Rides", title = "Number of Rides by Rider Type",fill="Rider Type")+
  scale_fill_manual(values = c("member" ='blueviolet','casual' = "darkorchid4"))

```
```{r, echo = FALSE}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()
  ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(x="Day of the Week", y="Average Duration", title = "Average Trip Duration by Rider Type",fill = "Rider Type")+
  scale_fill_manual(values = c("member" = "blueviolet", "casual" = "darkorchid4"))

```
## Exporting Summary File for Further Analysis
```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual +
all_trips_v2$day_of_week, FUN = mean)
write.csv(counts,"C:/Users/parke/OneDrive/Documents/R Codes/avg_ride_length.csv", row.names = FALSE)
write.csv(all_trips_v2,"C:/Users/parke/OneDrive/Documents/R Codes/all_trips_v2.csv")
```


